Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome

### 2017-08-08 docs/manuscript.rmd
updated

---
title: A putative causal map of natural variation across the human phenome
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
header-includes:
   - \usepackage{amsmath}
bibliography: manuscript.bib
csl: vancouver.csl
---

@@

A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data. MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype. Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a range of pleiotropy models, instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow. Methods have been developed in an attempt to protect MR from different patterns of horizontal pleiotropy, and here we have designed a mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates. Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, comprising the evaluation of causality between over a million pairs of traits. Almost 90% of causal estimates indicated some level of horizontal pleiotropy. The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)), and we use it here to highlight the numerous challenges that remain in automated causal inference.

@@

Genome-wide association studies (GWAS) have identified potential instrumental variables for thousands of phenotypes [@Hindorff2010]. Recent developments in MR have enabled knowledge of instrumental variables to be applied using either individual-level or only summary-level data (known as two-sample MR, 2SMR) [@Pierce2013]. Here, in order to infer the causal effect of an exposure on an outcome all that is required are the genetic effects of the instrumenting SNP on the exposure and the outcome. This has three major advantages. First, GWAS summary data are non-disclosive and often publicly available. Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting possible MR analyses only by the availability of GWAS summary data for the traits in question [@Hemani2016]. Third, because each instrumental variable mimics an independent randomization study, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis.

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure (Box 1). The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017]. It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017]. However there are circumstances in which it is desirable to consolidate across many methods to obtain a single MR estimate. First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient. Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no HP, it can have high false positive or low true positive rates in the presence of HP compared to other methods. Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g. horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known. Fourth, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.

@@

Pleiotropy is the phenomenon whereby a single genetic variant influences multiple phenotypes. The mechanisms behind it can arise through many different mechanisms, for example a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016]. How we classify the pleiotropic mechanism depends on the application, and in this paper we discuss pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait. It has also been termed, in some contexts, Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998]. It is this form of pleiotropy that MR assumes to make causal inference between two traits.

@@

In its simplest implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome. This is known as the inverse-variance weighted (IVW) method.

@@

**Mean-based methods:** Here we consider four nested models [@Bowden2017]. The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no HP. IVW random effects meta-analysis relaxes the HP assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression line without affecting the slope (and therefore not introducing bias). Fixed effects Egger regression [@Bowden2015] relaxes the HP assumption further by allowing a non-zero intercept which essentially allows overall horizontal pleiotropy to be directional, where its total effect influences the outcome in a specific direction. Random effects Egger regression further allows heterogeneity around the slope having accounted for overall directional HP [@Bowden2017], as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].

@@

The four nested models (IVW fixed effects, IVW random effects, Egger fixed effects, Egger random effects) plus the three Rucker estimates (point estimate, mean of the jackknife, median of the jackknife) provide seven mean based estimators.

**Median-based methods:** An alternative approach is to take the median effect of all available instruments [@Han2008; @Bowden2016b]. This has the advantage that only half the instruments need to be valid, and the estimate will remain unbiased. The weighted median estimate develops the approach further to allow stronger instruments to contribute more towards the estimate can be obtained by weighting the contribution of each instrument by the inverse of its variance. The penalised weighted median estimator introduces a further weight to the instruments, penalising any instrument that contributes substantially towards the heterogeneity statistic. Together, this provides three median based estimators. Other estimation strategies not considered here, such as LASSO regression, have also been developed for the situation where at least half of the instruments are valid [@Kang2016].

**Mode-based methods:** The mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017]. There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME). The simple mode is the unweighted mode of the empirical density function of causal estimates, whereas the weighted mode is weighted by the inverse variance of the outcome effect. The bandwidth parameter was set to 1 by default.

@@

The simplest approach to selecting instruments is to take SNPs that have been declared significant in the published GWAS for the exposure. This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample. These results are often recorded in public GWAS catalogs. Alternatively the clumping procedure can be performed using complete summary data in MR-Base [@Hemani2016]. Complete summary data refers to association results for all SNPs used in a GWAS i.e. not only those passing statistical thresholds for significance. We call this the "tophits" strategy.

@@

With genome-wide association studies growing ever larger, the statistical power to detect significant associations that may be influencing the trait downstream of many other pathways increases. For example, if a SNP $g_{A}$ influences trait $A$, and trait $A$ influences trait $B$, then a sufficiently powered GWAS will identify the $g_{A}$ as being significant for trait $B$ (Figure 1a). Using $g_{A}$ as an instrument to test the causal effect of $A$ on $B$ is perfectly valid. But the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$ could erroneously result in the apparent causal effect of $B$ on $A$. If $g_{A}$ is only one of many known instruments for $B$, amongst which some are valid, it is to the advantage of the researcher to exclude $g_{A}$ from the analysis.

@@

because the $cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}$. This simple inequality will not hold in some cases, for example $\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}$ where $\rho_{x, x_o}$ and $\rho_{y, y_o}$ are the precision of the measurements of the $x$ and $y$. Some combinations of confounding effects can also distort the $\rho_{g,x}$ and $\rho_{g,y}$ parameters, as has been discussed in detail previously [@Hemani2017]. However, we use it here as a computationally inexpensive and approximate method to identify variants that are likely to be invalid. Steiger's Z-test of correlated correlations [@Steiger1980] can be used to formally test the extent to which the two correlations are statistically different.

Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a). Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are more likely to be a source of HP. A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones. This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], or the *SLC2A9* variant influencing serum urate levels [@white2016plasma].

@@

We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total (Supplementary table 2). Each method assumes or performs best for a different model of pleiotropy. Our objective is to select the method most likely to be correct for a specific MR analysis by predicting the model of pleiotropy, and relating that prediction to the most appropriate model. To achieve this we use a "mixture of experts" machine learning approach [@Jordan1994], where each MR method is considered to be an 'expert', taking a summary-set as its input. An overview of the approach is shown in Figure 2.

@@

It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation. In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data. As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem. From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.

@@

1. If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method. This value is chosen arbitrarily as a minimum number of variants for which the different MR methods can model HP
2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
3. If there is 1 variant retained in the Steiger approach then use the Wald ratio on the remaining variant
4. If there are no variants remaining after Steiger filtering then declare no estimate of a causal association possible.

For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported. The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org).

@@

As statistical power for GWAS studies improves the likelihood of a significant association being discovered for an outcome trait that acts primarily on some exposure trait for that outcome increases (Figure 1a). We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated summary-sets comprising both null and non-null causal models. For each summary-set, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.

Figure 1b shows that the tophits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype. For brevity, we refer to the latter type of invalid instruments as “reverse causal” instruments. The proportion of invalid instruments due to reverse cause increased as GWAS discovery sample size increased. Applying Steiger filtering reduced this to 25%. Consequently, the false discovery rates (FDR) for 12 of the 14 methods reduced substantially when applied using Steiger filtered instruments (Figure 1c). The true positive rates for the methods based on Steiger filtering did however reduce slightly for 10 of the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular summary-set based on its characteristics (Figure 2).

@@

We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods. Figures 3a and 3b show that the MoE approach had the best general performance. Estimating the area under the receiver operator curve gave 0.84 for the MoE approach in terms of classifying the simulations as being null or non-null. Notably, the next best methods were median and mode based estimators using Steiger filtering. But a crucial observation is that under the assumption of widespread and diverse pleiotropic effects it is clear that all methods suffer from high false discovery rates (Figure 3a), including the MoE approach.

### Automated MR analysis of 2407 phenotypes

@@

The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3. Amongst those deemed 'significant', the IVW fixed effects analysis method combined with the tophits instruments selection strategy (only valid when there is no detectable evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases. This indicates that horizontal pleiotropy is likely to be pervasive.

### Putative associations involving LDL cholesterol and years of schooling

As an example of the database we performed a look up of associations between all traits and LDL cholesterol (as measured in the GLGC consortium [@Willer2013]), where the MoE result had a false discovery rate of 0.05 using the MoE approach. This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol. A large proportion of these traits were metabolites, which are dominated by lipid fractions. Filtering to exclude the metabolomic studies ([@Shin2014; @Kettunen2016]) returned 27 associations with 23 traits (Figure 4a). There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$). Here the IVW random effects method was chosen after Steiger filtering, indicating the presence of HP being present amongst the instruments.

We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences. Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b). Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017]. However, it is clear that there are serious limitations in relying on MR as a panacea for causal inference. For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g. childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible. Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits due to a third unmeasured trait, an intuitive interpretation of causality cannot be applied because the genetic instruments used proxy the liability to achieve some level of educational attainment, and not necessarily the experience of having attained a particular level of edication.

The facility to perform these scans has been automated in the MR of everything vs everything (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).

@@

Methods have been introduced recently that seek to improve MR estimates through invalid instrument detection and removal [@Corbin2016; @Verbanck2017; @Zhu2017]. These attack the problem from the perspective that variants which have a relatively large influence on the causal estimate, or provide substantially different estimates from the other variants, are likely to be exhibiting HP, and their removal from the analysis may be warranted. A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones. This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], because CRP itself is downstream of many cellular functions and most GWAS signals will not be specific for the trait. Steiger filtering has attempted to sidestep this issue in removing invalid instruments, seeking to detect those instruments which have associations with the outcome that are too large to be consistent with the exposure causing the outcome. We acknowledge that unmeasured confounding and measurement error can introduce problems with Steiger filtering [@Hemani2017], but note that our simulations indicate that all individual MR methods improve in performance following its application.

In addition to HP, there are many other limitations that prevent MR from being a panacea for causal inference. Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes. Such associations can be explained statistically because the instruments proxy the liability for that particular phenotype, not that phenotype's observed manifestation. Though these associations can be informative, their interpretations as causal relationships are far from clear. Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS of disease is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease. Again, interpretation of such associations can be problematic due to the instruments proxying disease liability and not diseae outcomes.

@@

Figure 2: Mixture of experts. a) Training. Datasets are simulated that have either null or non-null causal relationships, and causal estimates are obtained from each of the 28 MR methods considered in this study. In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations. The columns in red are specific to each method, they represent how well that method performed in obtaining the correct answer for each of the datasets. Random forests are used to learn the parameter space of the 53 metrics in which a particular dataset is likely to perform well. Together, this creates 28 random forest decision trees, one for each method. b) Application. For a GWAS summary dataset, our objective is to choose the method most likely to return the correct causal estimate. Metrics are generated from the dataset and fed into each of the 28 random forest decision trees. This provides us with 28 performance predictions. Finally, we use the method for which the performance prediction is highest.

>>>>

---
title: Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
header-includes:
   - \usepackage{amsmath}
bibliography: manuscript.bib
csl: vancouver.csl
---

@@

A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data. MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype. Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a range of pleiotropy models, instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow. Methods have been developed in an attempt to protect MR from different patterns of horizontal pleiotropy, and here we have designed a mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates. Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes. Almost 90% of causal estimates indicated some level of horizontal pleiotropy. The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)), and we use it here to highlight the numerous challenges that remain in automated causal inference.

@@

Genome-wide association studies (GWAS) have identified potential instrumental variables for thousands of phenotypes [@Hindorff2010]. Recent developments in MR have enabled knowledge of instrumental variables to be applied using either individual-level or only summary-level data (known as two-sample MR, 2SMR) [@Pierce2013]. Here, in order to infer the causal effect of an exposure on an outcome all that is required are the genetic effects of the instrumenting SNP on the exposure and the outcome. This has three major advantages. First, GWAS summary data are non-disclosive and often publicly available. Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting possible MR analyses only by the availability of GWAS summary data for the traits in question [@Hemani2016]. Third, because each instrumental variable mimics an independent randomization study, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis [@Bowden2017].

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy, where they influence the outcome through a pathway other than the exposure (Box 1). The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of horizontal pleiotropy [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017]. It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017]. However there are circumstances in which it is desirable to consolidate across many methods to obtain a single MR estimate. First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient. Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no horizontal pleiotropy, it can have high false positive or low true positive rates in the presence of horizontal pleiotropy compared to other methods. Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g. horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, if different methods disagree it is not possible to know which is correct because the true nature of horizontal pleiotropy exhibited by the instruments is not known. Fourth, the available methods do not cover all possible models of horizontal pleiotropy, and therefore an automated method for instrument selection may be necessary.

@@

Pleiotropy is the phenomenon whereby a single genetic variant influences multiple phenotypes. The mechanisms behind it can arise through many different mechanisms, for example a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016]. How we classify the pleiotropic mechanism depends on question and context, and in this paper we discuss pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait. It has also been termed, in some contexts, Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998]. It is this form of pleiotropy that MR assumes to make causal inference between two traits.

@@

In its simplest implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome. This is known as the IVW method.

@@

**Mean-based methods:** Here we consider four nested models [@Bowden2017]. The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no horizontal pleiotropy. IVW random effects meta-analysis relaxes the horizontal pleiotropy assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression line without affecting the slope (and therefore not introducing bias). Fixed effects Egger regression [@Bowden2015] relaxes the horizontal pleiotropy assumption further by allowing a non-zero intercept which essentially allows overall horizontal pleiotropy to be directional, where its total effect influences the outcome in a specific direction. Random effects Egger regression further allows heterogeneity around the slope having accounted for overall directional horizontal pleiotropy [@Bowden2017], as long as the horizontal pleiotropy effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].

@@

The four nested models (IVW fixed effects, IVW random effects, Egger fixed effects, Egger random effects) plus the three Rucker estimates (point estimate, mean of the jackknife, median of the jackknife) provide seven mean-based estimators.

**Median-based methods:** An alternative approach is to take the median effect of all available instruments [@Han2008; @Bowden2016b]. This has the advantage that only half the instruments need to be valid, and the estimate will remain unbiased. The weighted median estimate develops the approach further to allow stronger instruments to contribute more towards the estimate can be obtained by weighting the contribution of each instrument by the inverse of its variance. The penalised weighted median estimator introduces a further weight to the instruments, penalising any instrument that contributes substantially towards the heterogeneity statistic. Together, this provides three median-based estimators. Other estimation strategies not considered here, such as LASSO regression, have also been developed for the situation where at least half of the instruments are valid [@Kang2016].

**Mode-based methods:** The mode-based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017]. There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME). The simple mode is the unweighted mode of the empirical density function of causal estimates, whereas the weighted mode is weighted by the inverse variance of the outcome effect. The bandwidth parameter was set to 1 by default.

@@

The simplest approach to selecting instruments is to take SNPs that have been declared significant in the published GWAS for the exposure. This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample. These results are often recorded in public GWAS catalogs. Alternatively the clumping procedure can be performed using complete summary data in MR-Base [@Hemani2016]. Complete summary data refers to association results for all SNPs used in a GWAS i.e. not only those passing statistical thresholds for significance. We call this the "top-hits" strategy.

@@

With genome-wide association studies growing ever larger, the statistical power to detect significant associations that may be influencing the trait downstream of many other pathways increases. For example, if a SNP $g_{A}$ influences trait $A$, and trait $A$ influences trait $B$, then a sufficiently powered GWAS will identify the $g_{A}$ as being significant for trait $B$ (Figure 1a). Using $g_{A}$ as an instrument to test the causal effect of $A$ on $B$ is perfectly valid in some cases. But the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$, for example, could erroneously result in the apparent causal effect of $B$ on $A$. It is to the advantage of the researcher to exclude $g_{A}$ from the analysis from the analysis of $B$ against $A$.

@@

because the $cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}$. This simple inequality will not hold in some cases, for example $cor(x, x_o) < cor(x,y)cor(y,y_o)$ where $cor(x, x_o)$ and $cor(y, y_o)$ are the precision of the measurements of the $x$ and $y$. Some combinations of confounding effects can also distort the $cor(g,x)$ and $cor(g,y)$ parameters, as has been discussed in detail previously [@Hemani2017]. Large differences in sample sizes between the exposure and outcome GWASs may also have a practical impact on the efficacy of this approach. However, we use it here as a computationally inexpensive and approximate method to identify variants that are likely to be invalid (Figure 1a). Steiger's Z-test of correlated correlations [@Steiger1980] can be used to formally test the extent to which the two correlations are statistically different.

Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are more likely to be a source of horizontal pleiotropy. A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones. This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], or the *SLC2A9* variant influencing serum urate levels [@white2016plasma].

@@

We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 strategies in total (Supplementary table 2). Each method assumes or performs best for a different model of pleiotropy. Our objective is to select the method most likely to be correct for a specific MR analysis by predicting the model of pleiotropy, and relating that prediction to the most appropriate model. To achieve this we use a "mixture of experts" machine learning approach [@Jordan1994], where each MR strategy is considered to be an 'expert', taking a summary-set as its input. An overview of the approach is shown in Figure 2.

@@

It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation. In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data. As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem. From this perspective the 28 MR strategies are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.

@@

1. If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method. This value is chosen arbitrarily as a minimum number of variants for which the different MR methods can model horizontal pleiotropy
2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
3. If there is 1 variant retained in the Steiger approach then use the Wald ratio on the remaining variant
4. If there are no variants remaining after Steiger filtering then declare no estimate of a causal association possible.

The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org). For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.

@@

As statistical power for GWAS studies improves the likelihood of a significant association being discovered for a trait that acts primarily on one of its precursors increases (Figure 1a). This presents a problem when GWAS significance is used as the sole criterion for instrument selection if the hypothesis being tested is either a) the trait causing the precursor, or b) the trait causing some other trait and the precursor causally relates to both (in which case the precursor is a confounder). We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated summary-sets comprising both null and non-null causal models. For each summary-set, instruments were selected based on the top-hits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.

Figure 1b shows that the top-hits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype. For brevity, we refer to the latter type of invalid instruments as “reverse causal” instruments. The proportion of invalid instruments due to reverse cause increased as GWAS discovery sample size increased. Applying Steiger filtering reduced this to 25%. Consequently, the false discovery rates (FDR) for 12 of the 14 methods reduced substantially when applied using Steiger filtered instruments (Figure 1c). The true positive rates for the methods based on Steiger filtering did however reduce slightly for 10 of the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 14 MR methods and two instrument selection strategies (giving a total of 28 MR strategies) to apply to a particular summary-set based on its characteristics (Figure 2).

@@

We compared the MoE performance in the simulations against each of the 28 strategies, testing to see if it outperformed all other single strategies. Figures 3a and 3b show that the MoE approach had the best general performance. Estimating the area under the receiver operator curve gave 0.84 for the MoE approach in terms of classifying the simulations as being null or non-null. Notably, the next best methods were median and mode based estimators using Steiger filtering. But a crucial observation is that under the assumption of widespread and diverse pleiotropic effects it is clear that all methods suffer from high false discovery rates (Figure 3a), including the MoE approach.

### Automated MR analysis of 2407 phenotypes reveals substantial horizontal pleiotropy

@@

The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3. Amongst those deemed 'significant', the IVW fixed effects analysis method combined with the top-hits instruments selection strategy (only valid when there is no detectable evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases. This indicates that horizontal pleiotropy is likely to be pervasive.

### Interpretation of the putative associations in a broad scan of phenotypes requires detailed follow up

We performed a look up of associations between all traits and LDL cholesterol (as measured in the GLGC consortium [@Willer2013]), where the result had a false discovery rate of 0.05 using the MoE approach. This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol. A large proportion of these traits were metabolites, which are dominated by lipid fractions. Filtering to exclude the metabolomic studies ([@Shin2014; @Kettunen2016]) returned 27 associations with 23 traits (Figure 4a). There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$). Here the IVW random effects method was chosen after Steiger filtering, indicating the presence of horizontal pleiotropy being present amongst the instruments.

We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences. Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b). Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017]. There is also a protective effect of more years of schooling on lower risk of myocardial infarction, which has been supported by detailed follow up analyses [] However, it is clear that there are limitations in relying on MR as a panacea for causal inference. For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g. childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible. Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits due to a third unmeasured trait, an intuitive interpretation of causality cannot be applied because the genetic instruments used proxy the liability to achieve some level of educational attainment, and not necessarily the experience of having attained a particular level of edication.

The facility to perform these scans has been automated in the MR of 'everything vs everything' (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).

@@

Methods have been introduced recently that seek to improve MR estimates through invalid instrument detection and removal [@Corbin2016; @Verbanck2017; @Zhu2017]. These attack the problem from the perspective that variants which have a relatively large influence on the causal estimate, or provide substantially different estimates from the other variants, are likely to be exhibiting horizontal pleiotropy, and their removal from the analysis may be warranted. A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones. This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], because CRP itself is downstream of many cellular functions and most GWAS signals will not be specific for the trait. Steiger filtering has attempted to sidestep this issue in removing invalid instruments, seeking to detect those instruments which have associations with the outcome that are too large to be consistent with the exposure causing the outcome. We acknowledge that unmeasured confounding and measurement error can introduce problems with Steiger filtering [@Hemani2017], but note that our simulations indicate that all individual MR methods improve in performance following its application.

In addition to horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference. Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes. Such associations can be explained statistically because the instruments proxy the liability for that particular phenotype, not that phenotype's observed manifestation. Though these associations can be informative, their interpretations as causal relationships are far from clear. Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS of disease is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease. Again, interpretation of such associations can be problematic due to the instruments proxying disease liability and not diseae outcomes.

@@

Figure 2: Mixture of experts. a) Training. Datasets are simulated that have either null or non-null causal relationships, and causal estimates are obtained from each of the 28 MR strategies considered in this study. In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations. The columns in red are specific to each strategy, they represent how well that strategy performed in obtaining the correct answer for each of the datasets. Random forests are used to learn the parameter space of the 53 metrics in which a particular summary-set is likely to perform well. Together, this creates 28 random forest decision trees, one for each MR strategy. b) Application. For a GWAS summary dataset, our objective is to choose the strategy most likely to return the correct causal estimate. Metrics are generated from the dataset and fed into each of the 28 random forest decision trees. This provides us with 28 performance predictions. Finally, we use the strategy for which the performance prediction is highest.
